---
title: "Introduction"
author: 
- name: Jiefei Wang
  affiliation: Roswell Park Comprehensive Cancer Center, Buffalo, NY
date: "`r Sys.Date()`"
output:
    BiocStyle::html_document:
        toc: true
        toc_float: true
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(AltStats)
```

ALTREP is a newly introduced feature in R since R 3.5. It is an alternative way to represent R's atomic vector. By defining a set of functions in C level, users are able to define their own ALTREP class and create an ALTREP object that behaves exactly the same as R's atomic vector at the R level. Please refer to [ALTREP overview](https://blog.revolutionanalytics.com/2017/09/altrep-preview.html) and [ALTREP C++ implementation](https://purrple.cat/blog/2018/10/14/altrep-and-cpp/) to see details about ALTREP.

While ALTREP is a very promising and ambitious improvement, due to the limited resources of R's core team, there still exists some problems with ALTREP at the time of this vignette is writting. One of the most significant one is that many R basic functions are not capable to deal with an ALTREP object when the object does not define a data pointer function. In other word, if an ALTREP object does not allocate its data in a contiguous block of memory and cannot provide a data pointer, most R's function will aware of this issue and throw an error. Although the problem has been fixing since ALTREP is released, it is still not a completely solved problem. This package provide a quick patch for this issue by defining S3 and S4 classes for different ALTREP types. 

In this vignette, we will use a simple ALTREP object to show how to use the package. We will mainly focus on S3 ALTREP object, because S4 class does not fully work with ALTREP in some R versions. First we create an ALTREP object and compare it with a regular R object

```{r}
## Regular R object
A <- runif(4)
## ALTREP object
## This function is defined in the package
## and not exported
B <- AltStats:::makeExampleAltrep(A)

## compare A and B
typeof(A)
typeof(B)

class(A)
class(B)



.Internal(inspect(A))
.Internal(inspect(B))

## Print their values
## No error to print A
print(A)
## will cause an error.
tryCatch(print(B), error=function(e) print(e))

## An alternative way to print the values
A[1:10]
B[1:10]
```
The variable `A` and `B` behave exactly the same, but actually the variable `B` is an ALTREP object. If your are using R 3.6, You will see that the default print function does not work with the ALTREP variable `B`, we have to use the subset operator `[` to print out the vectors. If you are using R 4.0, you will not see the error since this problem has been fixed.

`print` function is not the only one that is not compatible with ALTREP. Here we will show a portion of them. It is possible that there is no error in your vignette, which means these problem has been fixed. There is no need to use the package.
```{r}
tryCatch(B + 1, error=function(e) print(e))
tryCatch(B == 1, error=function(e) print(e))
tryCatch(abs(B), error=function(e) print(e))
tryCatch(cumsum(B), error=function(e) print(e))
```

To solve this issue, the package provides S3 and S4 classes to override the default R functions. You can create an S3 or S4 object by calling `newAltWrapper`. 
```{r}
C <- newAltWrapper(B, S3Class = TRUE, forceSet = TRUE)
```
Note that we use the argument `forceSet` to force R to set the class attribute on the variable `B` since R cannot duplicate the object B and cause an error. This argument violates R's copy-on-change philosophy and also changes the variable `B`, so it should not be used unless you do not have any other options. If you are using R 4.0 you do not need this argument. Now let's check the variable `C`
```{r}
C

## Note that the variable B is also changed
## because forceSet = TRUE
## Highly unexpected!
B


C + 1
C == 1
abs(C)
cumsum(C)
```
The variable `C` works as usual, most functions in `?"group generic"` has been overrided to support ALTREP.

#Session info
```{r}
sessionInfo()
```














